---
title: "Getting Started with ML-SAS"
author: "Woohyung Kim and Sang-Il Lee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ML-SAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>")
library(mlsas)
```

## A small complete analysis

This vignette uses 36 synthetic
regions, 20 trees and 199 draws for speed. These are explicit teaching settings,
not evidence of adequate inferential precision. The defaults are 1,000 trees,
99,999 draws, path proximity, total global and conditional local randomization,
two-sided inference and BH adjustment at 0.05.

```{r}
d <- mlsas_example("mixed")
str(d$attributes)
fit <- mlsas(d$data, variables=d$variables, id=d$id,
             ntree=20, nperm=199, seed=2026,
             verbose=FALSE, progress=FALSE)
summary(fit)
head(as.data.frame(fit$local))
```

```{r map, fig.width=6, fig.height=5}
plot(fit)
```

Significant statistics above the conditional permutation mean are similarity
clusters; those below it are dissimilarity/outliers. These labels are not
univariate High-High or Low-Low categories. Non-significance is not proof of absence.
With B draws the minimum one-sided Monte Carlo p-value is 1/(B+1), and the minimum
for the implemented doubled-tail two-sided rule is 2/(B+1). BH adjustment may
require substantially more draws than a short example provides.

## Three input routes

An sf object contains attributes and geometry. Select analysis variables explicitly;
IDs and geometry never become forest attributes. Separate tables are joined by ID,
not by row order. ID sets must match exactly; no row is silently dropped.

```{r}
fit_separate <- mlsas(d$attributes, d$variables, d$id,
  spatial=d$spatial[nrow(d$spatial):1,], ntree=20, nperm=199,
  seed=2026, verbose=FALSE, progress=FALSE)
stopifnot(identical(fit$similarity, fit_separate$similarity) ||
          isTRUE(all.equal(unname(as.vector(fit$similarity)),
                           unname(as.vector(fit_separate$similarity)))))
w_rook <- mlsas_weights(d$data, id=d$id, contiguity="rook")
fit_weights <- mlsas(d$attributes, d$variables, d$id, weights=w_rook,
  ntree=20, nperm=199, seed=2026, verbose=FALSE, progress=FALSE)
```

Attributes may be a data.frame or tibble. Numeric columns are continuous;
factor columns are nominal; ordered factors retain their declared order. Convert
character categories explicitly. Missing or constant attributes are rejected.
Nominal factors are limited to 31 levels in this release. Geometry must be valid.

## Reuse the learned representation

```{r}
s <- fit$similarity
o <- mlsas_stat(s, w_rook)
upper <- mlsas_local_test(o, alternative="greater", nperm=199,
                         seed=8, verbose=FALSE, progress=FALSE)
head(as.data.frame(upper))
```

The forest is not refitted during permutation inference. Global inference jointly
relabels rows and columns; local conditional inference fixes the focal profile.
Total local inference is available with `randomization="total"`. The two schemes
answer different randomization questions. Adjustment applies to the current set
of input regions, not across multiple fitted specifications.

## Spatial choices and execution

Queen and rook require polygons. Distance and nearest-neighbour weights require
POINT sf geometry with a known CRS; explicitly choose representative points for
polygons. Distance thresholds use `sf::st_distance` units. Nearest-neighbour
symmetrization is `union`, `mutual` or `none`. Isolated regions produce an error.
Use `mlsas_weights` first for these advanced choices. Supplied weights are not
silently standardized again by the integrated function.

`verbose=TRUE` announces stages and measured time. `progress=interactive()` shows
completed tree/permutation work in interactive R. Forest fitting reports running;
there is no invented percent or ETA. `workers=2` parallelizes proximity calculation,
not fitting or inference. Use one worker when process spawning is unavailable.
A supplied seed restores the caller's RNG state after the call. The integrated
function consumes a single stream through fitting and both tests; separate calls
with their own seeds are not expected to reproduce that exact stream automatically.

## Resources and paper reproduction

Similarity storage is quadratic in region count. Tests stream 256-draw batches;
`keep=TRUE` stores the full distribution and uses additional memory. A fixed seed
alone does not guarantee equality across engines, languages or package versions.
The paper's full simulation campaign ran on AWS. This small example does not
reproduce that campaign. See the repository reproduction documentation for full
settings, archived runtimes and completed validation scope.
